Local PCA Shows How the Effect of Population Structure Differs Along the Genome

### 2018-10-23 writeup/review-responses.tex

\vskip 2em

\noindent
{\bf To the Editor(s) -- }

\vskip 1em

We are pleased to submit a revised version of our manuscript, 
``Local PCA shows how the effect of population structure differs along the genome''.

@@

\vskip 4em

\pagebreak
\setcounter{page}{1}

%%%%%%%%%%%%%%
\reviewersection{AE}

\begin{quote}
     Please include a response to each of the reviewers' comments. It is most
    important that you address the following in a revised manuscript: Reviewer 1's
    second comment; and more strongly addressing reviewer 2's concern on the issue
    of the number of PCs included would be helpful as the concern remains. Finally,
    I add some minor comments, and the senior editor has suggested other
    discussions in the literature to link to (extensive discussion is not needed).
\end{quote}

\begin{point}{Line 17.}
    ``meso-scale'' is vague (and more typically restricted to the physical domain?) - perhaps omit or replace
\end{point}

\reply{
    Good point; we've replaced it with ``intermediate-scale'' (maybe better?). \revref
}

\begin{point}{Line 44.}
    After ``partially reproductively isolated'' (e.g. due to partial post-zygotic incompatibilities) to make it more accessible/clear to a reader following the argument? I stumbled at first reading ``random mating'' and ``partially reproductively isolated'' but that apparent contradiction is exactly the point!
\end{point}

\reply{
    Another good point; we've got the word ``post-zygotic'' in there now. \revref
}

\begin{point}{Line 66.}
    Minor wording : PCA on the genetic data matrix (getting the genetic covariance matrix is an intermediate step in some algorithms); And Menozzi et al used frequency data, so maybe it's best to say inspired by pioneering work by Menozzi et al?
\end{point}

\reply{
    We agree, that is more accurate. Thanks; fixed. \revref
}

\begin{point}{Line 395--396} ``correlations [...] should not extend no further than does linkage disequilibrium'' - I get the point, but it's a tautology as written
\end{point}

\reply{
    We've clarified this to ``correlations in patterns of relatedness'',
    which we think isn't tautological any more,
    if LD is correlations in allelic identity, anyhow? \revref
}

\begin{point}{Figures S5 and S6:} 
    Include a panel which gives a legend/image to explain the colors being used for the bottom PCA such that the reader can know what it will look like when one of the PCAs in that row is mirroring geography.
\end{point}

\reply{
    Done.
}

\begin{point}{Senior editor comments:}
    It would be helpful to expand the discussion of variation in relationships
    along the genome - as written, the discussion is strongly focused on the
    "classical" Drosophila-centred linked selection literature.  Specifically,
    there has been a lot of discussion in the phylogenetics world about how to
    estimate species trees from the set of locus-specific genealogies (where the
    "consensus" tree can be quite wrong); and there are methods that use the
    distribution of genealogical relationships to estimate population structure
    (Konrad Lohse, Asger Hobolth etc).  It would be helpful to make these broader
    links.  
\end{point}

\reply{
}

@@

%%%%%%%%%%%%%%
\reviewersection{1}

\begin{quote}
    Li \& Ralph in their manuscript show how local PCA can be used to visualise genomic regions that differ from each other in the mean relatedness among samples. While a similar approach has be previously used to detect inversions in human genome (Ma & Amos 2012), they have applied their method to also other than human datasets and invoke other interpretations to the observed patterns than inversions.

    I appreciate the quality of writing (readability, clarity, organisation) and also the time spent amending the manuscript after the first review. The figures were clear and demonstrated the points they should. The method was very well explained and the code is understandable to use and easily accessible. The manuscript was fun to read.

    I have not previously reviewed the manuscript and but I think that most of the comments of previous reviewers were well reacted upon. However, the resulting manuscript is to a new reader a bit unclear in terms of the application and usability of the method and ability to distinguish between different scenarios leading to observed patterns and I think further simulations of mechanisms of interest should be performed with the aim to either demonstrate what the limitations of the method are for potential users or to further explore biological processes hinted at.
\end{quote}

\begin{point}{}
    Is the method meant to facilitate discovery of new inversions? If so, what number of samples are necessary? Of what coverage (what level of missingness is allowed)? What size of inversions is possible to detect? What are ranges of the distance values between windows that could point to inversions (any significance test would be recommended)? How does the method compare to other methods that detect inversions (esp. Ma & Amos 2012)? If it is not meant to discover new inversions and it should only confirm known existing inversions, what is the success rate (with varying sample size, missingness)?
\end{point}

\reply{
}

\begin{point}{}
    The authors argue that linked selection is the likely cause for the patterns they observed in Drosophila dataset since the recombination rate correlates with the position of the first MDS coordinate (from line number 321). At the same time, the authors mention that the ability to identify the windows could be influenced by recombination rate in the case of linked selection and also in the presence of recombination hotspots (section 3.1 Validation). As such, I do not agree with the authors, that it is enough to demonstrate that linked selection could cause these patterns but to use this in interpretation, they should be able to distinguish this explanation from the other possible causes in simulations first and then on real data (e.g. using recombination maps for Drosophila to simulate exactly the same regions could be of advantage).
\end{point}

\reply{
}

\begin{point}{}
    The authors refer to supplementary Figures sometimes as "Figure S6" and sometimes as "Supplementary Figure S6". In certain journals, these two could denote different figures so I would select one way only.
\end{point}

\reply{
}

\begin{point}{}
    I do not understand while when plotting the positions, the authors opted to show 2 graphs (for each MDS coordinate) when identifying outliers along the genome. While it is connected to the MDS plot they also show, I would rather try to express this in 1 value because I think that it would increase their ability to detect the outliers (e.g. Figure S6 - the 2 top left figures showing results of simulations with constant recombination rate when compared to MDS coordinate 1 and 2 - if the information from the two graphs was combined in 1 measure, I would expect the pattern to be more visible).
\end{point}

\reply{
}

@@

%%%%%%%%%%%%%%
\reviewersection{2}

\begin{quote}
    The authors have put in a considerable effort to address the concerns of the reviewers. The simulation studies are now much better and the conclusions were weakened somewhat in light of them. The method definitely appears to be useful. I think that it could be an important part of the literature that moves us forward in the empirical understanding of how the genome evolves.
\end{quote}

\begin{point}{}
I only have one major comment, which is unfortunate because it could have been avoided if the simulation were implemented slightly differently. The authors didn't address the intention of my concern about the switching of PCs. To be very explicit: when you perform a PC analysis, there is some variability about the order that the "true" directions of the data comes out in terms of the magnitude of that PC (i.e. the order). This model uses K PCs for each small segment of genome, and each small segment is compared using a method called MDS. It is convincing that order doesn't affect comparisons of the segments *provided that all of the components are present*. However, if the magnitude of the noise is high, then it is quite plausible that the top K components in a segment of genome are not the top K for the whole genome. The simulation is useless for reassuring on this because it sets up a situation with 2 important PCs which are always included, and the claim that "the simulation also verifies insensitivity to ordering of top PCs" is not really true. The real data analyses are more reassuring, given the relatively similar results between K=2 and K=5, but I think the concern remains valid.

I would like to see evidence that the windows that are found by this method are not associated with "which" PCs happen to be included in the top K. To do that, you could for example calculate the top M>K pcs, check the correlation with the overall PCs, and then plot "presence" of a PC (measured in correlation in the top K) against the MDS values. It would alternatively be possible to make an argument that it will always be fine if you use a large enough number of PCs, but this is argued to have severe computational problems (an argument I don't know that I buy). It would be enough to do this for retaining *one* PC in the simulation that has 2 *equally strong* PCs, but I have a suspicion that this case goes quite wrong and is anyway unrealistic. Better to keep more, e.g. 5 when there is non-trivial structure in the truth, and see that the 4th or 5th swapping with the 6th or 7th doesn't affect much. To be clear: the authors shouldn't necessarily follow this route, but they should address the underlying concern.
\end{point}

\reply{
}

\begin{point}{}
The simulation with geographically linked selection is not comparable to the simulation with 1/30 beneficial selection and 29/30 negative selection, because the probability of any mutation being positive of negative is now 15/30 (which depends on where in the geography the allele is). This makes this scenario a factor 15 higher in positive selection strength. The scenarios are ok, but this imbalance likely explains the difference, as positive selection is more likely to stand out than negative selection? 
\end{point}

\reply{
}

### 2018-10-23 writeup/local_pca_paper.tex
biorxiv latex
start at responses

% Responses to reviews
\input{review-response-commands}
% set this to show line numbers and include responses to reviews or not
\newif\ifreviewresponses
\reviewresponsestrue  % include them
% \reviewresponsesfalse  % don't include them
\newcommand{\responsefile}{review-responses.tex}  % name of the review reponses file

@@

@@

\ifsubmission\linenumbers\fi

@@

Population structure leads to systematic patterns in measures of mean relatedness
between individuals in large genomic datasets,
which are often discovered and visualized using dimension reduction techniques such as principal component analysis (PCA).
Mean relatedness
is an average of the relationships across locus-specific genealogical trees,
which can be strongly affected on intermediate genomic scales by linked selection and other factors.
We show how to use local principal components analysis to describe this 
meso-scale heterogeneity in patterns of relatedness,
and apply the method to genomic data from three species,
finding in each that the effect of population structure 
can vary substantially across only a few megabases.
In a global human dataset, localized heterogeneity is likely explained by polymorphic chromosomal inversions.
In a range-wide dataset of \textit{Medicago truncatula},
factors that produce heterogeneity are shared between chromosomes,
correlate with local gene density,
and may be caused by linked selection, such as background selection or local adaptation.
In a dataset of primarily African \textit{Drosophila melanogaster}, large-scale heterogeneity across each chromosome arm
is explained by known chromosomal inversions thought to be under recent selection,
and after removing samples carrying inversions, remaining heterogeneity is correlated with recombination rate and gene density,
again suggesting a role for linked selection.
% The method is implemented as an R package, called lostruct.
The visualization method provides a flexible new way to discover biological drivers of genetic variation,
and its application to data highlights the strong effects that linked selection and chromosomal inversions
can have on observed patterns of genetic variation.

@@

\citetx[Wright]{wright1949genetical} defined \emph{population structure} to encompass ``such matters as numbers, composition by age and sex, and state of subdivision'',
where ``subdivision'' refers to restricted migration between subpopulations.
The phrase is also commonly used to refer to the genetic patterns that result from this process,
as for instance reduced mean relatedness between individuals from distinct populations.
However, it is not necessarily clear what aspects of demography should be included in the concept. \revpoint{2}{10}
For instance, 
\citetx[Blair]{blair1943population} defines \emph{population structure} to be the sum total of
``such factors as size of breeding populations, periodic fluctuation of population size, sex ratio, activity range and \emph{differential survival of progeny}'' (emphasis added).
The definition is similar to Wright's, but differs in including the effects of natural selection.
On closer examination, incorporating differential survival or fecundity makes the concept less clear:
should a randomly mating population consisting of two types that are partially reproductively isolated from each other
be said to show population structure or not?
Whatever the definition,
it is clear that due to natural selection, the effects of population structure --
the \emph{realized} patterns of genetic relatedness --
differ depending on which portion of the genome is being considered.
For instance, strongly locally adapted alleles of a gene will be selected against in migrants to different habitats,
increasing genetic differentiation between populations near to this gene.
Similarly, newly adaptive alleles spread first in local populations.
These observations motivate many methods to search for genetic loci under selection,
as for example in \citet{huertasnchez2013genetic}, \citet{martin2016natural}, 
and \citet{duforetfrebourg2015detecting}.

@@

Patterns in genome-wide relatedness are often summarized
by applying principal components analysis (PCA, \citet{patterson2006population}) 
to the genetic covariance matrix,
as pioneered by \citet{menozzi1978synthetic}.
% About 37 years ago, \citet{menozzi1978synthetic} first applied principal component analysis (PCA) in population genetics to construct maps summarizing genetic variation \citep{menozzi1978synthetic}. 
% Nowadays, PCA is a widely used powerful non-parametric method to extract information from genetic data.
The results of PCA can be related to the genealogical history of the samples, 
such as time to most recent common ancestor and migration rate between populations \citep{novembre2008interpreting,mcvean2009genealogical}, 
and sometimes produce ``maps'' of population structure
that reflect the samples' geographic origin distorted by rates of gene flow
\citep{novembre2008genes}.
% although other patterns can be caused by recent migration or nongeographic kinship patterns.  % \citep{astle2009population}.
% Through dimension-reduction, PCA can identify key components of population structure, which describes how different samples are related, and are often closely related to geography. 
% Plots of the first two principal components (PCs) can mimic the samples' geographic origin to some extent. 
% Since population structure describes how different samples are related, samples living closer tend to be more genetically similar and thus tend to be clustered in PC plots \citep{novembre2008genes,patterson2006population}. 
% However this relatedness is limited while there's recent migration or for group with nongeographic kinship patterns,for example, social or religious groups. \citep{astle2009population}

@@

Others have applied PCA to windows of the genome:
\citet{ma2012investigation} used local PCA much as we do 
to identify putative chromosomal inversions.
\citet{bryc2010genomewide} and \citet{brisbin2012pcadmix}
used PCA to infer tracts of local ancestry in recently admixed populations,
but by projecting each genomic window onto the axes of a single, globally-defined PCA
rather than doing PCA separately on each window.  \revpoint{1}{7}

@@

It is common to describe variation along the genome of simple statistics such as $F_{ST}$
and to interpret the results in terms of the action of selection 
(e.g., \citet{turner2005genomic,ellegren2012genomic}).
However, a given pattern (e.g., valleys of $F_{ST}$)
can be caused by more than one biological process \citep{cruickshank2014reanalysis,burri2015linked},
which in retrospect is unsurprising given that we are using a single statistic
to describe a complex process.
It is also common to use methods such as PCA to visualize large-scale patterns in 
mean genome-wide relatedness.
In this paper we show if and how patterns of mean relatedness vary systematically along the genome, 
in a way particularly suited to large samples from geographically distributed populations.
Geographic population structure sets the stage by establishing ``background'' patterns of relatedness;
our method then describes how this structure is affected by selection and other factors.
Our aim is not to identify outlier loci,
but rather to describe larger-scale variation shared by many parts of the genome;
correlation of this variation with known genomic features
can then be used to uncover its source. \revpoint{2}{1}

@@

The top 2--5 principal components are generally good summaries of population structure; 
for ease of visualization we usually only use the first two (referred to as $\pcone$ and $\pctwo$),
and check that results hold using more. \revpoint{1}{6}
The above procedure can be performed on any subset of the data;
for future reference, denote by $\pcone_j$ and $\pctwo_j$
the result after applying to all SNPs in the $j^\text{th}$ window.
(Note, however, that our measure of dissimilarity between windows
does not depend on PC ordering.) \llabel{ll:another_pc_point}

@@

We think of the local effects of population structure as being summarized by \emph{relative} position of the samples
in the space defined by the top principal components.
However, 
we do not compare patterns of relatedness of different genomic regions by directly comparing the PCs,
since rotations or reflections of these imply identical patterns of relatedness.
Instead, we compare the low-dimensional approximations of the local covariance matrices
obtained using the top $k$ PCs,
which is invariant under reordering of the PCs, \revpoint{1}{2}
reflections, and rotations and yet contains all other information about the PCs.
(For results shown here, we use $k=2$.)
Furthermore, to remove the effect of artifacts such as mutation rate variation,
we also rescale each approximate covariance matrix to be of similar size
(precisely, so that the underlying data matrix has trace norm equal to one).

@@

We applied the method to windows of average length 9 Kbp, across chromosome arms 2L, 2R, 3L, 3R and X separately.
The first column of \Figure \ref{fig:mds_allchr} is a multidimensional scaling (MDS) visualization 
of the matrix of dissimilarities between genomic windows:
in other words, genomic windows that are closer to each other in the MDS plot show more similar patterns of relatedness.
For each chromosome arm, the MDS visualization roughly resembles a triangle,
sometimes with additional points.
Since the relative position of each window in this plot shows the similarity between windows, 
this suggests that there are at least three extreme manifestations of population structure 
typified by windows found in the ``corners'' of the figure,
and that other windows' patterns of relatedness may be a mixture of those extremes. 
The next two columns of \Figure \ref{fig:mds_allchr} respectively depict the two MDS coordinates of each window,
plotted against the window's position along the genome,
to show how the plot of the first column is laid out along the genome.
The patterns did not depend on the number of PCs used
(see \Figure \ref{sfig:dmel_k5} for the same plot with $k=5$ PCs), \revpoint{1}{3}
and are only weakly correlated with variation in missingness 
(see \Figure \ref{sfig:dpgp_missingness}). \revpoint{2}{4}

@@

Unlike the other two species,
the method applied separately on all eight chromosomes of \textit{Medicago truncatula} 
showed similar patterns of gradual change in patterns of relatedness across each chromosome,
with no indications of chromosome-specific patterns.
This consistency suggests that the factor affecting the population structure for each chromosome is the same,
as might be caused by varying strengths of linked selection.
To verify that variation in the effects of population structure is shared across chromosomes,
we applied the method to all chromosomes together.
Results for chromosome 3 are shown in \Figure \ref{fig:mds12_medicago}, \revpoint{2}{13}
and other chromosomes are similar:
across chromosomes, the high values of the first MDS coordinate coincide with the position of the heterochromatic regions surrounding the centromere,
which often have lower gene density and may therefore be less subject to linked selection.
% Selection on genes in Medicago has been demonstrated by \citet{paape2013selection}. 
To verify that this is a possible explanation,
we counted the number of genes found in each window using gene models in Mt4.0 from \url{jcvi.org} \citep{tang2014improved},
which are shown juxtaposed with
the first MDS coordinate of each window in \Figure \ref{fig:mds_medicago},
and are significantly correlated, as shown in Supplemental \Figure \ref{sfig:mds_gene_count}.
(Values shown are the number of start and end positions of each predicted mRNA transcript,
divided by two, assigned to the nearest window.)
% that lie closer to SNPs  between the last SNP of each window and the first SNP in the next window.
However, other genomic features, such as distance to centromere show roughly the same patterns,
so we cannot rule out alternative hypotheses.
In particular, fine-scale recombination rate estimates are not available in a form mappable
to Mt4.0 coordinates (although those in \citet{paape2012finescale} appear visually similar).
\revpoint{2}{8}

The results were highly consistent across window sizes, window types (SNPs or bp),
and number of PCs, as shown in Table \ref{tab:param_cors}.
\revpoint{1}{4} \revpoint{1}{5}
% We also found nearly identical results when choosing shorter windows of 1,000 SNPs
% (Supplemental Figure \ref{sfig:mds_medicago_allchr_1e3}; \revpoint{1}{4}
% or choosing windows of equal length in base pairs rather than SNPs. \revpoint{1}{5}
% Similarly, the results were not substantially changed
% when using weighted PCA to downweight the large group of Tunisian samples.

@@

\paragraph{The effect of selection}
Neutral processes are not expected to produce the chromosome-scale correlations we see in
patterns of relatedness 
in the \textit{Medicago truncatula} and \textit{Drosophila melanogaster} datasets, \llabel{ll:not_neutral}
because correlations induced by neutral processes 
should extend no further than does linkage disequilibrium
(i.e., much less than a chromosome's length).
This suggests that they are produced by linked selection,
a hypothesis backed up by correlations with gene density and recombination rate.
We have also shown with simulations that linked selection can, in at least some circumstances,
produce the sorts of patterns we observe.
How might selection cause variation in patterns of relatedness?
For instance, background selection
(the effect on linked sites of selection against deleterious mutations
\citet{charlesworth1993effect,charlesworth2013background}),
can informally be thought of as reducing the number of potential contributors to the gene pool 
in regions of the genome with many possible deleterious mutations \citep{hudson1995deleterious}.
For this reason, if it acts in a spatial context, it is expected to induce samples from nearby locations to cluster together more frequently.
Therefore, regions of the genome harboring many targets of local adaptation may show similar patterns,
since migrant alleles in these regions will be selected against,
and so locally gene trees will more closely reflect spatial proximity.
Other forms of selection, such as hard sweeps on new mutations,
repeated selection on standing variation, local adaptation,
or temporally fluctuating selection,
could clearly lead to variation 
in geographic patterns of relatedness in a similar way.

@@

There has been substantial debate over the relative impacts of different forms of selection
(e.g., \citet{charlesworth1997effects,charlesworth2012effects,pease2013accurate,hedrick2013adaptive,burri2015linked,corbettdetig2015natural,harris2016genetic,martin2016natural,phung2016determining,stankowski2018tempo}). \revpoint{2}{16}
These have been difficult to disentangle in part because for the most part
theory makes predictions which are only strictly valid in randomly mating (i.e., unstructured) populations,
and it is unclear to what extent the spatial structure observed in most real populations will affect these predictions.
It may be possible to design more powerful statistics that make stronger use of spatial information.

@@

\paragraph{Applications}
So-called cryptic relatedness between samples
has been one of the major sources of confounding in genome-wide association studies (GWAS)
and so methods must account for it by modeling population structure or kinship \citep{astle2009population,yang2014advantages}.
Modern ``mixed model'' methods \citep[e.g.][]{loh2015efficient} 
account for this with either a single, genome-wide kinship matrix
or one constructed using only sites unlinked to the focal SNP. \revpoint{2}{9}   %unlike ---> unlink?
Since the effects of population structure is not constant along the genome,
this could in principle lead to an inflation of false positives in parts of the genome
with stronger population structure than the genome-wide average.
A method such as ours might be used to estimate local kinship matrices,
thus providing a more sensitive correction,
although doing so without removing the signal itself could be challenging.
Fortunately, in our human dataset this does not seem likely to have a strong effect:
most variation is due to small, independent regions, possibly primarily inversions,
and so may not have a major effect on GWAS.
In the other species we examined, particularly \textit{Drosophila melanogaster},
treating population structure as a single quantity would entail a substantial loss of power,
and could potentially be misleading.

>>>>

% Responses to reviews
\input{review-response-commands}
\newif\ifbiorxiv
\biorxivfalse
% \biorxivtrue
\ifbiorxiv
    \renewcommand{\revpoint}[2]{\relax}
    \renewcommand{\llabel}[1]{\relax}
\fi
% set this to show line numbers and include responses to reviews or not
\newif\ifreviewresponses
\reviewresponsestrue  % include them
% \reviewresponsesfalse  % don't include them
\ifbiorxiv\reviewresponsesfalse\fi
\newcommand{\responsefile}{review-responses.tex}  % name of the review reponses file

@@

\ifbiorxiv\nolinenumbers\fi

@@

@@

Population structure leads to systematic patterns in measures of mean relatedness
between individuals in large genomic datasets,
which are often discovered and visualized using dimension reduction techniques such as principal component analysis (PCA).
Mean relatedness
is an average of the relationships across locus-specific genealogical trees,
which can be strongly affected on intermediate genomic scales by linked selection and other factors.
We show how to use local principal components analysis to describe this 
intermediate-scale heterogeneity in patterns of relatedness, \revpoint{AE}{1}
and apply the method to genomic data from three species,
finding in each that the effect of population structure 
can vary substantially across only a few megabases.
In a global human dataset, localized heterogeneity is likely explained by polymorphic chromosomal inversions.
In a range-wide dataset of \textit{Medicago truncatula},
factors that produce heterogeneity are shared between chromosomes,
correlate with local gene density,
and may be caused by linked selection, such as background selection or local adaptation.
In a dataset of primarily African \textit{Drosophila melanogaster}, large-scale heterogeneity across each chromosome arm
is explained by known chromosomal inversions thought to be under recent selection,
and after removing samples carrying inversions, remaining heterogeneity is correlated with recombination rate and gene density,
again suggesting a role for linked selection.
% The method is implemented as an R package, called lostruct.
The visualization method provides a flexible new way to discover biological drivers of genetic variation,
and its application to data highlights the strong effects that linked selection and chromosomal inversions
can have on observed patterns of genetic variation.

@@

\citetx[Wright]{wright1949genetical} defined \emph{population structure} to encompass ``such matters as numbers, composition by age and sex, and state of subdivision'',
where ``subdivision'' refers to restricted migration between subpopulations.
The phrase is also commonly used to refer to the genetic patterns that result from this process,
as for instance reduced mean relatedness between individuals from distinct populations.
However, it is not necessarily clear what aspects of demography should be included in the concept. % \revpoint{2}{10}
For instance, 
\citetx[Blair]{blair1943population} defines \emph{population structure} to be the sum total of
``such factors as size of breeding populations, periodic fluctuation of population size, sex ratio, activity range and \emph{differential survival of progeny}'' (emphasis added).
The definition is similar to Wright's, but differs in including the effects of natural selection.
On closer examination, incorporating differential survival or fecundity makes the concept less clear:
should a randomly mating population consisting of two types 
that exhibit partial post-zygotic reproductive isolation from each other \revpoint{AE}{2}
be said to show population structure or not?
Whatever the definition,
it is clear that due to natural selection, the effects of population structure --
the \emph{realized} patterns of genetic relatedness --
differ depending on which portion of the genome is being considered.
For instance, strongly locally adapted alleles of a gene will be selected against in migrants to different habitats,
increasing genetic differentiation between populations near to this gene.
Similarly, newly adaptive alleles spread first in local populations.
These observations motivate many methods to search for genetic loci under selection,
as for example in \citet{huertasnchez2013genetic}, \citet{martin2016natural}, 
and \citet{duforetfrebourg2015detecting}.

@@

Patterns in genome-wide relatedness are often summarized
by applying principal components analysis (PCA, \citet{patterson2006population}) 
to the genetic data matrix,
as inspired by the pioneering work of \citet{menozzi1978synthetic}. \revpoint{AE}{3}
% About 37 years ago, \citet{menozzi1978synthetic} first applied principal component analysis (PCA) in population genetics to construct maps summarizing genetic variation \citep{menozzi1978synthetic}. 
% Nowadays, PCA is a widely used powerful non-parametric method to extract information from genetic data.
The results of PCA can be related to the genealogical history of the samples, 
such as time to most recent common ancestor and migration rate between populations \citep{novembre2008interpreting,mcvean2009genealogical}, 
and sometimes produce ``maps'' of population structure
that reflect the samples' geographic origin distorted by rates of gene flow
\citep{novembre2008genes}.
% although other patterns can be caused by recent migration or nongeographic kinship patterns.  % \citep{astle2009population}.
% Through dimension-reduction, PCA can identify key components of population structure, which describes how different samples are related, and are often closely related to geography. 
% Plots of the first two principal components (PCs) can mimic the samples' geographic origin to some extent. 
% Since population structure describes how different samples are related, samples living closer tend to be more genetically similar and thus tend to be clustered in PC plots \citep{novembre2008genes,patterson2006population}. 
% However this relatedness is limited while there's recent migration or for group with nongeographic kinship patterns,for example, social or religious groups. \citep{astle2009population}

@@

Others have applied PCA to windows of the genome:
\citet{ma2012investigation} used local PCA much as we do 
to identify putative chromosomal inversions.
\citet{bryc2010genomewide} and \citet{brisbin2012pcadmix}
used PCA to infer tracts of local ancestry in recently admixed populations,
but by projecting each genomic window onto the axes of a single, globally-defined PCA
rather than doing PCA separately on each window.  % \revpoint{1}{7}

@@

It is common to describe variation along the genome of simple statistics such as $F_{ST}$
and to interpret the results in terms of the action of selection 
(e.g., \citet{turner2005genomic,ellegren2012genomic}).
However, a given pattern (e.g., valleys of $F_{ST}$)
can be caused by more than one biological process \citep{cruickshank2014reanalysis,burri2015linked},
which in retrospect is unsurprising given that we are using a single statistic
to describe a complex process.
It is also common to use methods such as PCA to visualize large-scale patterns in 
mean genome-wide relatedness.
In this paper we show if and how patterns of mean relatedness vary systematically along the genome, 
in a way particularly suited to large samples from geographically distributed populations.
Geographic population structure sets the stage by establishing ``background'' patterns of relatedness;
our method then describes how this structure is affected by selection and other factors.
Our aim is not to identify outlier loci,
but rather to describe larger-scale variation shared by many parts of the genome;
correlation of this variation with known genomic features
can then be used to uncover its source. % \revpoint{2}{1}

@@

The top 2--5 principal components are generally good summaries of population structure; 
for ease of visualization we usually only use the first two (referred to as $\pcone$ and $\pctwo$),
and check that results hold using more. % \revpoint{1}{6}
The above procedure can be performed on any subset of the data;
for future reference, denote by $\pcone_j$ and $\pctwo_j$
the result after applying to all SNPs in the $j^\text{th}$ window.
(Note, however, that our measure of dissimilarity between windows
does not depend on PC ordering.) \llabel{ll:another_pc_point}

@@

We think of the local effects of population structure as being summarized by \emph{relative} position of the samples
in the space defined by the top principal components.
However, 
we do not compare patterns of relatedness of different genomic regions by directly comparing the PCs,
since rotations or reflections of these imply identical patterns of relatedness.
Instead, we compare the low-dimensional approximations of the local covariance matrices
obtained using the top $k$ PCs,
which is invariant under reordering of the PCs, % \revpoint{1}{2}
reflections, and rotations and yet contains all other information about the PCs.
(For results shown here, we use $k=2$.)
Furthermore, to remove the effect of artifacts such as mutation rate variation,
we also rescale each approximate covariance matrix to be of similar size
(precisely, so that the underlying data matrix has trace norm equal to one).

@@

We applied the method to windows of average length 9 Kbp, across chromosome arms 2L, 2R, 3L, 3R and X separately.
The first column of \Figure \ref{fig:mds_allchr} is a multidimensional scaling (MDS) visualization 
of the matrix of dissimilarities between genomic windows:
in other words, genomic windows that are closer to each other in the MDS plot show more similar patterns of relatedness.
For each chromosome arm, the MDS visualization roughly resembles a triangle,
sometimes with additional points.
Since the relative position of each window in this plot shows the similarity between windows, 
this suggests that there are at least three extreme manifestations of population structure 
typified by windows found in the ``corners'' of the figure,
and that other windows' patterns of relatedness may be a mixture of those extremes. 
The next two columns of \Figure \ref{fig:mds_allchr} respectively depict the two MDS coordinates of each window,
plotted against the window's position along the genome,
to show how the plot of the first column is laid out along the genome.
The patterns did not depend on the number of PCs used
(see \Figure \ref{sfig:dmel_k5} for the same plot with $k=5$ PCs), % \revpoint{1}{3}
and are only weakly correlated with variation in missingness 
(see \Figure \ref{sfig:dpgp_missingness}). % \revpoint{2}{4}

@@

Unlike the other two species,
the method applied separately on all eight chromosomes of \textit{Medicago truncatula} 
showed similar patterns of gradual change in patterns of relatedness across each chromosome,
with no indications of chromosome-specific patterns.
This consistency suggests that the factor affecting the population structure for each chromosome is the same,
as might be caused by varying strengths of linked selection.
To verify that variation in the effects of population structure is shared across chromosomes,
we applied the method to all chromosomes together.
Results for chromosome 3 are shown in \Figure \ref{fig:mds12_medicago}, % \revpoint{2}{13}
and other chromosomes are similar:
across chromosomes, the high values of the first MDS coordinate coincide with the position of the heterochromatic regions surrounding the centromere,
which often have lower gene density and may therefore be less subject to linked selection.
% Selection on genes in Medicago has been demonstrated by \citet{paape2013selection}. 
To verify that this is a possible explanation,
we counted the number of genes found in each window using gene models in Mt4.0 from \url{jcvi.org} \citep{tang2014improved},
which are shown juxtaposed with
the first MDS coordinate of each window in \Figure \ref{fig:mds_medicago},
and are significantly correlated, as shown in Supplemental \Figure \ref{sfig:mds_gene_count}.
(Values shown are the number of start and end positions of each predicted mRNA transcript,
divided by two, assigned to the nearest window.)
% that lie closer to SNPs  between the last SNP of each window and the first SNP in the next window.
However, other genomic features, such as distance to centromere show roughly the same patterns,
so we cannot rule out alternative hypotheses.
In particular, fine-scale recombination rate estimates are not available in a form mappable
to Mt4.0 coordinates (although those in \citet{paape2012finescale} appear visually similar).
% \revpoint{2}{8}

The results were highly consistent across window sizes, window types (SNPs or bp),
and number of PCs, as shown in Table \ref{tab:param_cors}.
% \revpoint{1}{4} \revpoint{1}{5}
% We also found nearly identical results when choosing shorter windows of 1,000 SNPs
% (Supplemental Figure \ref{sfig:mds_medicago_allchr_1e3}; \revpoint{1}{4}
% or choosing windows of equal length in base pairs rather than SNPs. \revpoint{1}{5}
% Similarly, the results were not substantially changed
% when using weighted PCA to downweight the large group of Tunisian samples.

@@

\paragraph{The effect of selection}
Neutral processes are not expected to produce the chromosome-scale correlations we see in
patterns of relatedness 
in the \textit{Medicago truncatula} and \textit{Drosophila melanogaster} datasets, \llabel{ll:not_neutral}
because correlations in patterns of relatedness induced by neutral processes \revpoint{AE}{4}
should extend no further than does linkage disequilibrium
(i.e., much less than a chromosome's length).
This suggests that they are produced by linked selection,
a hypothesis backed up by correlations with gene density and recombination rate.
We have also shown with simulations that linked selection can, in at least some circumstances,
produce the sorts of patterns we observe.
How might selection cause variation in patterns of relatedness?
For instance, background selection
(the effect on linked sites of selection against deleterious mutations
\citet{charlesworth1993effect,charlesworth2013background}),
can informally be thought of as reducing the number of potential contributors to the gene pool 
in regions of the genome with many possible deleterious mutations \citep{hudson1995deleterious}.
For this reason, if it acts in a spatial context, it is expected to induce samples from nearby locations to cluster together more frequently.
Therefore, regions of the genome harboring many targets of local adaptation may show similar patterns,
since migrant alleles in these regions will be selected against,
and so locally gene trees will more closely reflect spatial proximity.
Other forms of selection, such as hard sweeps on new mutations,
repeated selection on standing variation, local adaptation,
or temporally fluctuating selection,
could clearly lead to variation 
in geographic patterns of relatedness in a similar way.

@@

There has been substantial debate over the relative impacts of different forms of selection
(e.g., \citet{charlesworth1997effects,charlesworth2012effects,pease2013accurate,hedrick2013adaptive,burri2015linked,corbettdetig2015natural,harris2016genetic,martin2016natural,phung2016determining,stankowski2018tempo}). % \revpoint{2}{16}
These have been difficult to disentangle in part because for the most part
theory makes predictions which are only strictly valid in randomly mating (i.e., unstructured) populations,
and it is unclear to what extent the spatial structure observed in most real populations will affect these predictions.
It may be possible to design more powerful statistics that make stronger use of spatial information.

@@

\paragraph{Applications}
So-called cryptic relatedness between samples
has been one of the major sources of confounding in genome-wide association studies (GWAS)
and so methods must account for it by modeling population structure or kinship \citep{astle2009population,yang2014advantages}.
Modern ``mixed model'' methods \citep[e.g.][]{loh2015efficient} 
account for this with either a single, genome-wide kinship matrix
or one constructed using only sites unlinked to the focal SNP. % \revpoint{2}{9}   %unlike ---> unlink?
Since the effects of population structure is not constant along the genome,
this could in principle lead to an inflation of false positives in parts of the genome
with stronger population structure than the genome-wide average.
A method such as ours might be used to estimate local kinship matrices,
thus providing a more sensitive correction,
although doing so without removing the signal itself could be challenging.
Fortunately, in our human dataset this does not seem likely to have a strong effect:
most variation is due to small, independent regions, possibly primarily inversions,
and so may not have a major effect on GWAS.
In the other species we examined, particularly \textit{Drosophila melanogaster},
treating population structure as a single quantity would entail a substantial loss of power,
and could potentially be misleading.
